Preparing your manuscript

### 2025-02-18 article/article.qmd
add meridional gradient of temperature change

global_map <- rnaturalearth::ne_coastline()
models_in_gadi <- c("CMCC-ESM2", "EC-Earth3", "INM-CM4-8", "INM-CM5-0", 
                    "MPI-ESM1-2-HR", "NorESM2-MM", "EC-Earth3-CC", "EC-Earth3-Veg",
                    "EC-Earth3-Veg-LR", "GFDL-CM4")
```

@@

sig <- data |> 
  _[(model %in% models_in_gadi) & is.finite(change)] |>
  dcast(lat + lon + model ~ var, value.var = "diff") |> 
  _[, .(sig = sum(sign(deltap97.5), na.rm = TRUE)), by = .(lon, lat)]

@@

mean_trend <- data |> 
  _[(model %in% models_in_gadi) & is.finite(change)] |>
  dcast(lat + lon + model ~ var, value.var = "diff") |> 
  _[, let(land = MaskLand(lon, lat))] |> 
  _[, let(land = factor(land, labels = c("Sea", "Land")))] |> 
  _[, .(mean_dt = mean(deltap97.5, na.rm = TRUE)), by = .(lat, land, model)] |> 
  ggplot(aes(mean_dt, lat)) +
  geom_hline(yintercept = 0, color = "grey80") +
  geom_vline(xintercept = 0, color = "grey80") +
  geom_path(aes(group = model), color = "grey70", alpha = 0.7) +
  stat_summary(orientation = "y", geom = "line") +
  scale_y_latitude(breaks = seq(-80, 80 , 20)) +
  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            fill = "#fbfbfb",
            data = annotation,
            inherit.aes = FALSE) +
  facet_wrap(~land) +
  tagger::tag_facets(tag_pool = c("b", "c")) +
  labs(x = "days/year") +
  coord_fixed(ratio = 1/2.8) +
  theme_minimal() +
  theme(legend.position = "bottom",
        panel.background = element_rect(fill = "#fbfbfb", color = NA),
        strip.text.x = element_text(size = 12),
        plot.margin = unit(c(0, 0 , 1.9, -0.40), units = "cm"),
        tagger.panel.tag.text = element_text(size = 12),
        tagger.panel.tag.background = element_rect(fill = "white", color = "white"))

@@

model_spread <- data |> 
  _[, let(land = MaskLand(lon, lat))] |> 
  _[, let(land = factor(land, labels = c("Sea", "Land")))] |> 
  _[, let(serie = fcase(lat %between% c(48, 50), "50 - NH",
                        lat %between% c(38, 40), "40 - NH",
                        lat %between% c(28, 30), "30 - NH", 
                        lat %between% c(-50, -48), "50 - SH",
                        lat %between% c(-40, -38), "40 - SH",
                        lat %between% c(-30, -28), "30 - SH"))] |>
  _[!is.na(serie) & var == "deltap97.5" & model %in% models_in_gadi] |>
  _[, .(mean_freq = weighted.mean(ssp585, cos(lat*pi/180))), by = .(serie, land, model)] |> 
  tidyr::separate(serie, into = c("lat", "hemisphere"), sep = " - ") |> 
  setDT() |> 
  _[, let(scenario = "SSP5-8.5")] |> 
  _[!(land == "Land" & hemisphere == "SH" & lat == 50)]

@@

```{r}
#| label: fig-eke
#| fig-width: 8
#| fig-height: 4
#| fig.cap: "Mean change in EKE and $\\Delta T$ frequency for SSP5-8.5 (2080-2100) compared with the historical period (1980-2000) for each model. Values are calculated for the northern (25$^{\\circ}$ to 65$^{\\circ}$) and southern (-25$^{\\circ}$ to -65$^{\\circ}$) hemispheres and over sea (circles) and land (triangles). Each solid line corresponds to a linear regression for each hemisphere. b) and c) show the zonal mean for EKE ($m^2 s^{-2}$) and $\\Delta T$ (days/year) over land and sea respectively"

change <- data |> 
  _[(model %in% models_in_gadi) & is.finite(change)] |>
  dcast(lat + lon + model ~ var, value.var = "diff") |>

  _[, let(land = MaskLand(lon, lat),
          region = fcase(lat %between% c(25, 65), "NH",
                         lat %between% c(-65, -25), "SH",
                         default = NA))] |> 
  _[, let(land = factor(land, labels = c("Sea", "Land")))] |> 
  na.omit() |> 
  _[, .(mean_eke = weighted.mean(eke, w = cos(lat*pi/180)),
        mean_deltat = weighted.mean(deltap97.5, w = cos(lat*pi/180))), by = .(region, land, model)] |> 
  ggplot(aes(mean_eke, mean_deltat)) +
  geom_point(aes(color = region, shape = land), size = 2) +
  geom_smooth(aes(colour = region), method = "lm", se = FALSE) +
  scale_color_manual(values = c("cyan4", "orange")) +
  labs(x = latex2exp::TeX("EKE change ($m^2 s^{-2}$)"), y = latex2exp::TeX("$\\Delta T$ frequency change (days/year)"),
       color = NULL, shape = NULL) +
  theme_minimal() +
  theme(legend.position = "inside",
        legend.position.inside = c(0.83, 0.08),
        legend.direction = "horizontal",
        legend.background = element_rect(fill = "#fbfbfb", color = "#fbfbfb"),
        legend.margin = margin(-3, 0, 0, 0, unit = "mm"),
        panel.background = element_rect(fill = "#fbfbfb", color = NA))

deltat_mean <- data |> 
  _[(model %in% models_in_gadi) & is.finite(change)] |>
  dcast(lat + lon + model ~ var, value.var = "diff") |> 
  _[, let(land = MaskLand(lon, lat))] |> 
  _[, let(land = factor(land, labels = c("Sea", "Land")))] |> 
  _[, .(mean_dt = mean(deltap97.5, na.rm = TRUE)), by = .(lat, land)]

mean_trend_dt <- data |> 
  _[(model %in% models_in_gadi) & is.finite(change)] |>
  dcast(lat + lon + model ~ var, value.var = "diff") |> 
  _[, let(land = MaskLand(lon, lat))] |> 
  _[, let(land = factor(land, labels = c("Sea", "Land")))] |> 
  _[, .(mean_eke = mean(eke, na.rm = TRUE)), by = .(lat, land, model)] |> 
  ggplot(aes(mean_eke, lat)) +
  geom_hline(yintercept = 0, color = "grey80") +
  geom_vline(xintercept = 0, color = "grey80") +
  geom_path(aes(group = model), color = "grey70", alpha = 0.7) +
  stat_summary(orientation = "y", geom = "line", aes(color = "EKE")) +
  geom_path(data = deltat_mean, aes(x = mean_dt, color = "DeltaT")) +
  scale_color_manual(values = c("DeltaT" = "red4", "EKE" = "black"),
                     labels = c(latex2exp::TeX("$\\Delta T$"), "EKE")) +
  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            fill = "#fbfbfb",
            data = annotation,
            inherit.aes = FALSE) +
  annotate(geom = "rect", ymin = c(-90, -20, 90), ymax = c(-65, 20, 65), xmin = -Inf, xmax = Inf,
           fill = "grey93", alpha = 0.7) +
  scale_y_latitude(breaks = seq(-80, 80 , 20)) +
  facet_wrap(~land) +
  tagger::tag_facets(tag_pool = c("b", "c")) +
  labs(x = "Change", color = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom",
        panel.background = element_rect(fill = "#fbfbfb", color = NA),
        strip.text.x = element_text(size = 12),
        plot.margin = unit(c(0, 0 , 0, -0.40), units = "cm"),
        tagger.panel.tag.text = element_text(size = 12))

plot_grid(change, mean_trend_dt, labels = c('a)', ''), 
          label_x = 0.10, label_y = 0.98,
          label_fontface = "plain", 
          label_colour = "black",
          rel_widths = c(1.75, 1))
```

>>>>

global_map <- rnaturalearth::ne_coastline()
models_in_gadi <- c("CMCC-ESM2", "EC-Earth3", "INM-CM4-8", "INM-CM5-0", 
                    "MPI-ESM1-2-HR", "NorESM2-MM", "EC-Earth3-CC", "EC-Earth3-Veg",
                    "EC-Earth3-Veg-LR", "GFDL-CM4")

tas <- purrr::map(Sys.glob(here("data/tas/*hist*.nc")), function(f) {

  message(f)
  meta <- unglue::unglue(basename(f), patterns = c("{var}_day_{model}_{scenario}_{member}_{grid}_20150101-21001231.nc"))

  file_scenario <- Sys.glob(here(paste0("data/", meta[[1]][["var"]], "/", 
                                        meta[[1]][["var"]], "_day_", meta[[1]][["model"]], "_ssp585*")))

  var <- meta[[1]][["var"]]

  ReadNetCDF(f, vars = c(hist = var)) |> 
    _[, let(ssp585 = ReadNetCDF(file_scenario, vars = c(ssp585 = var), out = "vector")[[1]])] |> 
    _[, let(model = meta[[1]][["model"]],
            var = "dt_dy",
            lat = round(lat, digits = 8),
            hist = metR:::.derv(hist, lat)*(180/pi/6371000),
            ssp585 = metR:::.derv(ssp585, lat)*(180/pi/6371000)), by = .(lon)] |> 
    _[, let(change = ssp585 / hist,
            diff = ssp585 - hist)]

data <- rbind(data, tas) |> 
  _[, let(land = MaskLand(lon, lat))] |> 
  _[, let(land = factor(land, labels = c("Sea", "Land")))]

@@

sig <- data |> 
  _[(model %in% models_in_gadi) & var == "deltap97.5"] |>
  dcast(lat + lon + model ~ var, value.var = "diff") |> 
  _[, .(sig = sum(sign(deltap97.5), na.rm = TRUE)), by = .(lon, lat)]

@@

mean_trend <- data |> 
  _[(model %in% models_in_gadi) & var == "deltap97.5"] |>
  dcast(lat + lon + model ~ var, value.var = "diff") |> 
  _[, let(land = MaskLand(lon, lat))] |> 
  _[, let(land = factor(land, labels = c("Sea", "Land")))] |> 
  _[, .(mean_dt = mean(deltap97.5, na.rm = TRUE)), by = .(lat, land, model)] |> 
  ggplot(aes(mean_dt, lat)) +
  geom_hline(yintercept = 0, color = "grey80") +
  geom_vline(xintercept = 0, color = "grey80") +
  geom_path(aes(group = model), color = "grey70", alpha = 0.7) +
  stat_summary(orientation = "y", geom = "line") +
  scale_y_latitude(breaks = seq(-80, 80 , 20)) +
  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            fill = "#fbfbfb",
            data = annotation,
            inherit.aes = FALSE) +
  facet_wrap(~land) +
  tagger::tag_facets(tag_pool = c("b", "c")) +
  labs(x = "days/year") +
  coord_fixed(ratio = 1/2.8) +
  theme_minimal() +
  theme(legend.position = "bottom",
        panel.background = element_rect(fill = "#fbfbfb", color = NA),
        strip.text.x = element_text(size = 12),
        plot.margin = unit(c(0, 0 , 1.9, -0.40), units = "cm"),
        tagger.panel.tag.text = element_text(size = 12),
        tagger.panel.tag.background = element_rect(fill = "white", color = "white"))

@@

model_spread <- data |> 
  _[, let(serie = fcase(lat %between% c(48, 50), "50 - NH",
                        lat %between% c(38, 40), "40 - NH",
                        lat %between% c(28, 30), "30 - NH", 
                        lat %between% c(-50, -48), "50 - SH",
                        lat %between% c(-40, -38), "40 - SH",
                        lat %between% c(-30, -28), "30 - SH"))] |>
  _[!is.na(serie) & var == "deltap97.5" & model %in% models_in_gadi] |>
  _[, .(mean_freq = weighted.mean(ssp585, cos(lat*pi/180))), by = .(serie, land, model)] |> 
  tidyr::separate(serie, into = c("lat", "hemisphere"), sep = " - ") |> 
  setDT() |> 
  _[, let(scenario = "SSP5-8.5")] |> 
  _[!(land == "Land" & hemisphere == "SH" & lat == 50)]

@@

```{r}
#| label: fig-eke
#| fig-width: 8
#| fig-height: 4
#| fig.cap: "Mean change in $\\Delta T$ frequency and a) $\\nabla _{y}$ and b) EKE for SSP5-8.5 (2080-2100) compared with the historical period (1980-2000) for each model. Values are calculated for the northern (25$^{\\circ}$ to 65$^{\\circ}$) and southern (-25$^{\\circ}$ to -65$^{\\circ}$) hemispheres and over sea (circles) and land (triangles). Each solid line corresponds to a linear regression for each hemisphere."

change <- data |> 
  _[(model %in% models_in_gadi)] |>
  dcast(lat + lon + model + land ~ var, value.var = "diff") |>

  _[, let(region = fcase(lat %between% c(25, 65), "NH",
                         lat %between% c(-65, -25), "SH",
                         default = NA))] |> 
  _[, .(mean_eke = weighted.mean(eke, w = cos(lat*pi/180), na.rm = TRUE),
        mean_dt_dy = weighted.mean(dt_dy, w = cos(lat*pi/180), na.rm = TRUE),
        mean_deltat = weighted.mean(deltap97.5, w = cos(lat*pi/180), na.rm = TRUE)), by = .(region, land, model)] |> 
  na.omit()

gradt <- change |> 
  _[, let(mean_dt_dy = fifelse(region == "NH", -mean_dt_dy, mean_dt_dy))] |> 
  ggplot(aes(mean_dt_dy*1000, mean_deltat)) +
  geom_point(aes(color = region, shape = land), size = 2) +
  geom_smooth(aes(colour = region), method = "lm", se = FALSE) +
  scale_color_manual(values = c("cyan4", "orange")) +
  labs(x = latex2exp::TeX("$\\nabla _{y} T$ change (K/km)"), y = latex2exp::TeX("$\\Delta T$ frequency change (days/year)"),
       color = NULL, shape = NULL) +
  theme_minimal() +
  theme(legend.position = "inside",
        legend.position.inside = c(0.83, 0.08),
        legend.direction = "horizontal",
        legend.background = element_rect(fill = "#fbfbfb", color = "#fbfbfb"),
        legend.margin = margin(-3, 0, 0, 0, unit = "mm"),
        panel.background = element_rect(fill = "#fbfbfb", color = NA))

eke <- change |> 
  ggplot(aes(mean_eke, mean_deltat)) +
  geom_point(aes(color = region, shape = land), size = 2) +
  geom_smooth(aes(colour = region), method = "lm", se = FALSE) +
  scale_color_manual(values = c("cyan4", "orange")) +
  labs(x = latex2exp::TeX("EKE change ($m^2 s^{-2}$)"), y = latex2exp::TeX("$\\Delta T$ frequency change (days/year)"),
       color = NULL, shape = NULL) +
  theme_minimal() +
  theme(legend.position = "inside",
        legend.position.inside = c(0.83, 0.08),
        legend.direction = "horizontal",
        legend.background = element_rect(fill = "#fbfbfb", color = "#fbfbfb"),
        legend.margin = margin(-3, 0, 0, 0, unit = "mm"),
        panel.background = element_rect(fill = "#fbfbfb", color = NA))

# deltat_mean <- data |> 
#   _[(model %in% models_in_gadi) & is.finite(change)] |>
#   dcast(lat + lon + model ~ var, value.var = "diff") |> 
#   _[, let(land = MaskLand(lon, lat))] |> 
#   _[, let(land = factor(land, labels = c("Sea", "Land")))] |> 
#   _[, .(mean_dt = mean(deltap97.5, na.rm = TRUE)), by = .(lat, land)]
# 
# mean_trend_dt <- data |> 
#   _[(model %in% models_in_gadi) & is.finite(change)] |>
#   dcast(lat + lon + model ~ var, value.var = "diff") |> 
#   _[, let(land = MaskLand(lon, lat))] |> 
#   _[, let(land = factor(land, labels = c("Sea", "Land")))] |> 
#   _[, .(mean_eke = mean(eke, na.rm = TRUE)), by = .(lat, land, model)] |> 
#   ggplot(aes(mean_eke, lat)) +
#   geom_hline(yintercept = 0, color = "grey80") +
#   geom_vline(xintercept = 0, color = "grey80") +
#   geom_path(aes(group = model), color = "grey70", alpha = 0.7) +
#   stat_summary(orientation = "y", geom = "line", aes(color = "EKE")) +
#   geom_path(data = deltat_mean, aes(x = mean_dt, color = "DeltaT")) +
#   scale_color_manual(values = c("DeltaT" = "red4", "EKE" = "black"),
#                      labels = c(latex2exp::TeX("$\\Delta T$"), "EKE")) +
#   geom_rect(aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
#             fill = "#fbfbfb",
#             data = annotation,
#             inherit.aes = FALSE) +
#   annotate(geom = "rect", ymin = c(-90, -20, 90), ymax = c(-65, 20, 65), xmin = -Inf, xmax = Inf,
#            fill = "grey93", alpha = 0.7) +
#   scale_y_latitude(breaks = seq(-80, 80 , 20)) +
#   facet_wrap(~land) +
#   tagger::tag_facets(tag_pool = c("b", "c")) +
#   labs(x = "Change", color = NULL) +
#   theme_minimal() +
#   theme(legend.position = "bottom",
#         panel.background = element_rect(fill = "#fbfbfb", color = NA),
#         strip.text.x = element_text(size = 12),
#         plot.margin = unit(c(0, 0 , 0, -0.40), units = "cm"),
#         tagger.panel.tag.text = element_text(size = 12))

plot_grid(gradt, eke, labels = c('a)', 'b)'), 
          label_x = 0.12, label_y = 0.98,
          label_fontface = "plain", 
          label_colour = "black",
          rel_widths = c(1, 1))
```
